Red Wine Exploration by Jose Ballesteros
========================================================
```{r init, include=FALSE}
knitr::opts_chunk$set(fig.width=9, fig.height=6,
                      echo=FALSE,warning=FALSE, message=FALSE)
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(corrplot)
library(dplyr)
library(corrplot)
library(gridExtra)
library(reshape2)
library(scales)

```

```{r}
# Load the Data
redWine <- read.csv("wineQualityReds.csv", sep = ",")

head(redWine)


```

```{r}
names(redWine)
```

> 
After loading the dataset, it was discovered that the X variable was
only used as an index. As a result, the X variable was removed.  


```{r}
#Removed the X variable as it was only an index variable. 
redWine1 <- subset( redWine, select = -X )
names(redWine1)
```

```{r}

dim(redWine1)

```

> 
The Red Wine dataset explores 11 physicochemical and sensory 
characteristics on 1,599 samples of Portuguese "Vinho Verde" 
Red Wine to their relation with overall quality. 

# Univariate Plots Section

```{r}
summary(redWine1)

```


```{r}
#fixed acidity plot
ggplot(data = redWine1, aes(x = fixed.acidity)) +
  geom_histogram() + xlab("Fixed Acidity (g/dm3)") + ylab("Count") 

```

> Fixed acidity measured the amount of tartaric acid in the red wine.
Fixed acidity ranged from 4.60 to 15.90 g/dm3. It has a positively 
screwed distribution with a median of 7.90 g/dm3.

```{r}
#Volatile Acidity plot
ggplot(data = redWine1, aes(x = volatile.acidity)) +
  geom_histogram() + ylab("Count") + 
  xlab("Volatile Acidity (g/dm3)") 

```

> Volatile Acidity measured the amount of acetic acid in the red wine
and ranged from .12 to 1.58 g/dm3. The median was .52 g/dm3. The 
distrution has a slight positive skewed output. 

```{r}
#Citric Acid Plot
ggplot(data = redWine1, aes(x = citric.acid)) +
  geom_histogram() + 
  ylab("Count") + 
  xlab("Citric Acid (g/dm3)") 

```

>  Citric acid has a positively skewed distribution with a range of 
0 to 1.0 g/dm3. The median had a value of .26 g/dm3.    

```{r}

#Residual Sugar Plot

ggplot(data = redWine1, aes(x = residual.sugar)) +
  geom_histogram()  + 
  scale_x_log10(breaks=seq(0,16,by=2)) + 
  ylab("Count") + xlab("Residual Sugar (g/dm3)")

```

> 
The residual sugar plot initially had a very high positively skewed 
distribution. As a result, the log10 was applied to residual sugar.The 
residual sugar ranged from .90 to 15.50 g/dm3. The median had a value 
of 2.20 g/dm3. 

```{r}
#Chlorides Plot
ggplot(data = redWine1, aes(x = chlorides)) +
  geom_histogram() + 
  scale_x_log10(breaks=seq(0,.70,by=.1)) + 
  ylab("Count") + xlab("Chlorides (g/dm3)")
```

> The amount of sodium chloride was measured to calculate the amount of 
chlorides in the different samples. The log10 was applied to reduce the
positive skewness it had before. The amount of chlorides ranged from 
.012 to .611 g/dm3 and had a median of .079 g/dm3. 

```{r}
#Free Sulfer dioxide
ggplot(data = redWine1, aes(x = free.sulfur.dioxide)) +
  geom_histogram() + ylab("Count") + 
  xlab("Free Sulfur Dioxide (mg/dm3)") 
```

>  
The free sulfur dioxide plot showed a high level of positive skewness.
The range was from 1.0 to 72.0 mg/dm3. The median had a value of 
14 mg/dm3.  

```{r}
#total sulfer dioxide
ggplot(data = redWine1, aes(x = total.sulfur.dioxide)) +
  geom_histogram() + scale_x_log10(breaks=seq(0,300,by=50)) +
  xlab("Total Sulfur Dioxide (mg/dm3)") +
  ylab("Count")
```

> 
The log10 was applied to the total sulfur dioxide plot. Without
the log10, it displayed a highly positive skewed distribution.The 
amount of total sulfur dioxide ranged from 6.0 to 289.0 mg/dm3.The
median value was of 38.0 mg/dm3. 


```{r}
#density
ggplot(data = redWine1, aes(x = density)) +
  geom_histogram() + ylab("Count") + 
  xlab("Density (g/cm3)")
```

> 
The density of the samples ranged from.9901 to 1.0037 g/cm3 and 
had a median value of .9968 g/cm3.This chart shows density to be
normally distributed. 


```{r}
#pH
ggplot(data = redWine1, aes(x = pH)) +
  geom_histogram() + ylab("Count")
```

> 
The pH ranged from 2.74 to 4.01 and had a median value of 3.310. 
The pH displays normal distribution. 

```{r}
#sulphates
ggplot(data = redWine1, aes(x = sulphates)) +
  geom_histogram() + ylab("Count") + 
  xlab("Sulphates (g/dm3)") + 
  scale_x_log10(breaks=seq(0,2,.2))
```

> 
Potassium sulphate was used to measure the amount of sulphates
in the red wine.The log10 was applied to reduce some of the 
positive skewness it had before. The amount of sulphates ranged 
from .33 to 2.00 g/dm3 and had median of .62 g/dm3. 

```{r}
#alcohol
ggplot(data = redWine1, aes(x = alcohol)) +
  geom_histogram() + ylab("Count") +
  xlab("Alcohol (% per volume)") 
```

> 
The alcohol content ranged from 8.40 to 14.90 percent per volume.
The median value was of 10.20. It shows a positively skewed 
distribution. 

```{r}
#quality
ggplot(data = redWine1, aes(x = as.factor(quality))) + 
  geom_bar() + ylab("Count") + xlab("Quality")
```

> 
Quality of the wines was the outcome of the study from this dataset.
The score ranged from 3 to 8. The median quality was 6. 

# Univariate Analysis

###  Dataset Structure
```{r}
str(redWine1)

```

> 
The structure of this dataset consisted of 11 input variables and 
quality as the output variable. All of the input variables were of 
numeric data types. The quality output variable was an integer value.
The quality variable was used throughout the remaining portion of this
analysis as a categorical variable with the help of the as.factor 
function. No new variables were created besides converting the data
type.

> 
My interest within the dataset was wanting to know more about what 
physiochemical properties would better describe the results of the
different quality scores. As shown in the quality histogram, the 
majority of the wines in the sample received a score of 5 and 6. 
Many univariate plots initially had a positively skewed plot. 
These plots were then modified by incorporating a log10 scale that
at times improved the distributions as is the case with chlorides,
residual sugars, sulphates and the total sulfur dioxide. The 
remaining plots that still have a slight positive skewness did not
change after applying the log10 modification. As a result, the 
log10 was not applied as there was no noticeable change. The 
density and pH plot were the only two that were normally distributed
without any further modifications. 

# Bivariate Plots Section

```{r}
Red <- cor(redWine1)
head(round(Red,2))
corrplot(Red, method = "number", type = "lower")

```

> 
The corrplot function was used to see which variables had high
correlation among them for further analysis.  

```{r}
ggplot(aes(factor(quality),
           alcohol),
           data = redWine1) +
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5,color = 'blue') +
  stat_summary(fun.y = "mean",
               geom = "point",
               color = 'red',
               shape = 8,
               size = 4) +
  labs(title="Comparison of quality and alcohol") +
  xlab("quality")


```

> 
The correlation between quality and alcohol was a positve relation 
at .4761.The jitter and alpha function were used to improve the
output of the plot. 

```{r}
cor.test(redWine1$quality, redWine1$alcohol)
```

```{r}
ggplot(data = redWine1, aes(x=alcohol, y=density)) +
  geom_point(alpha = 1/8) + 
  labs(title="Comparison of alcohol and density") +
  geom_smooth(method = "lm",color = 'red',se = FALSE)
```

> Alcohol and density had a negative relationship with a correlation
of -.496.The alpha function was used to improve the output of the 
scatter plot.  

```{r}
cor.test(redWine1$alcohol, redWine1$density)
```

```{r}
ggplot(data = redWine1, 
       aes(x=total.sulfur.dioxide,
           y=free.sulfur.dioxide)) +
  geom_point(alpha=1/10) + 
  labs(title=
         "Comparison of total sulfur dioxide and free sulfur dioxide")
```

> 
The total sulfur and free sulfur dioxide had a positive 
relationship with a correlation of .668. 

```{r}
cor.test(redWine1$total.sulfur.dioxide, 
         redWine1$free.sulfur.dioxide)
```

```{r}
ggplot(data = redWine1, aes(x=pH, y=citric.acid )) +
  geom_point(alpha=1/10) + 
  labs(title="Comparison of pH and citric acid")
```

> 
Citric acid and pH had a negative relation with a 
correlation of -.542. 

```{r}
cor.test(redWine1$pH, redWine1$citric.acid)
```

```{r}
ggplot(data = redWine1, aes(x=pH, y=fixed.acidity)) + 
  geom_point(alpha=1/10) + 
  labs(title="Comparison of pH and fixed acidity")
```

> Fixed acidity and pH had a negative relation with a correlation
of -.683.

```{r}
cor.test(redWine1$pH, redWine1$fixed.acidity)

```

```{r}
ggplot(data = redWine1, 
       aes(x=citric.acid, y=volatile.acidity )) +
  geom_point(alpha=1/6) + 
labs(title="Comparison of citric acid and volatile acidity")
```

> Citric acid and volatile acidity had a negative relation
with a correlation of -.552. 

```{r}
cor.test(redWine1$citric.acid, redWine1$volatile.acidity)
```

```{r}
ggplot(data = redWine1, aes(x=citric.acid, y=fixed.acidity)) + 
  geom_point(alpha=1/8) + 
  labs(title="Comparison of citric acid and fixed acidity")
```

> Citric acid and fixed acidity displays a positive relation with 
a correlation of .672.  

```{r}
cor.test(redWine1$citric.acid, redWine1$fixed.acidity)
```

```{r}
ggplot(data = redWine1, aes(x=density, y=fixed.acidity )) +
  geom_point(alpha=1/10) + 
  labs(title="Comparison of density and fixed acidity")
```

> Density and fixed acidity displays a positive relation with a
correlation of .668. 

```{r}
cor.test(redWine1$density, redWine1$fixed.acidity)
```

# Bivariate Analysis

> 
After seeing the corrplot, the strongest input variable that had
a moderate positive correlation with quality was alcohol. The 
remaining variables had very low positive correlations with quality.
Volatile acidity had the highest negative correlation with quality 
at -.39.There were also several variables that had strong positve 
correlations among each other. Fixed acidity had a positive 
correlation with both citric acid and density at .67. Another positive
correlation of .67 was also seen with total and free sulfur dioxide. 
A slightly lower positive correlation was also seen with alcohol and
quality at .48.The highest negative correlation was between fixed 
acidity and pH levels at -.68. This was followed by citric acid and
volatile acidity with a negative .55 value. This relationship was 
interesting as they both relate to the flavor that will characterize
the red wine to have. Citric acid also shared a negative correlation 
with pH at -.54. Finally, the last variable analyzed was density and 
alcohol at a negative correlation of .5. 

# Multivariate Plots Section


```{r}
ggplot(data = redWine1, aes(x=density, y=fixed.acidity )) +
  geom_point(aes(color = as.factor(quality))) + 
  scale_color_brewer(type = 'seq') + 
  labs(title="Comparison of density and fixed acidity") +
  labs(color = "Wine Quality") + 
  geom_smooth(method = "lm",color = 'red',se = FALSE) +
  theme(legend.position = "bottom") +
  theme_dark() + xlab("Density") +
  ylab("Fixed Acidity")
```

>  
The bulk of the wines had a quality score of 5 and 6 with a 
median value of 7.8 and .997 for the fixed acidity and density.
Higher quality wines of 7 and above had a median value of 8.7 
and .996 for the fixed acidity and density. The lower quality 
wines of 4 and 3 had a median of 7.5 for the fixed acidity and a 
density of .997. 


```{r}
t1 <- redWine1 %>% filter(quality >= 5 & quality <=6 ) %>% 
  select(fixed.acidity,density) %>%
  summarise(
    fixedAcidMedian = median(fixed.acidity),
    densityMedian = median(density),
    n = n()
  )
t2 <-  redWine1 %>% filter(quality >= 7) %>%
  select(fixed.acidity,density) %>%
  summarise(
    fixedAcidMedian = median(fixed.acidity),
    densityMedian = median(density),
    n = n())
t3 <-  redWine1 %>% filter(quality <= 4) %>% 
  select(fixed.acidity,density) %>%
  summarise(
    fixedAcidMedian = median(fixed.acidity),
    densityMedian = median(density),
    n = n())
knitr::kable(list(t1, t2,t3))
    
```

```{r}
#multivariate graph with fixed acidity, citric acid and quality

ggplot(data = redWine1, aes(x=citric.acid, y=fixed.acidity )) + 
  geom_point(aes( color = as.factor(quality))) +
  scale_color_brewer(type = 'seq') + 
  labs(title="Comparison of citric acid and fixed acidity") +
  labs(color = "Wine Quality") + 
  geom_smooth(method = "lm",color = 'red',se = FALSE) + 
  theme(legend.position = "bottom") +
  theme_dark() + 
  xlab("Citric Acid") + 
  ylab("Fixed Acidity")

```

>  
The majority of the redwines of quality 5 and 6 are centered 
around .24 for the citric acid and 7.8 for fixed acidity. The 
higher quality wines are around .4 for the citric acid content 
and 8.7 for fixed acidity. The lower quality wines are centered 
around .08 for citric acid and 7.5 for fixed acidity. 

```{r}
a1 <- redWine1 %>% filter(quality >= 5 & quality <=6 ) %>% 
  select(fixed.acidity,citric.acid) %>%
  summarise(
    fixedAcidMedian = median(fixed.acidity),
    citricAcidMedian = median(citric.acid),
    n = n()
  )
a2 <-  redWine1 %>% filter(quality >= 7) %>% 
  select(fixed.acidity,citric.acid) %>%
  summarise(
    fixedAcidMedian = median(fixed.acidity),
    citricAcidMedian = median(citric.acid),
    n = n())
a3 <-  redWine1 %>% filter(quality <= 4) %>% 
  select(fixed.acidity,citric.acid) %>%
  summarise(
    fixedAcidMedian = median(fixed.acidity),
    citricAcidMedian = median(citric.acid),
    n = n())
knitr::kable(list(a1, a2,a3))
```


```{r}
#multivariate graph with volatile acidity, citric acid and quality

ggplot(data = redWine1, aes(x=citric.acid, y=volatile.acidity )) +
  geom_point(aes(color = as.factor(quality)))  + 
  scale_color_brewer(type = 'seq') + 
  labs(title=
"Comparison of citric acid and volatile acidity in terms of quality") + 
  labs(color = "Wine Quality")+ 
  geom_smooth(method = "lm",color = 'red',se = FALSE) +
  theme(legend.position = "bottom") +
  theme_dark() + xlab("Citric Acid") + 
  ylab("Volatile Acidity")

```

> 
The plot shows the majority of the wines with a quality of 5 
centered above the line of best fit while wines of quality of 
6,7, and 8 predominately are found below the line of best fit.
As citric acid is increased and volatile acidity is reduced the
quality tends to improve. 

```{r}

#multivariate graph with pH, fixed acidity and quality

ggplot(data = redWine1, aes(x=pH, y=fixed.acidity )) + 
  geom_point(aes(color = as.factor(quality))) + 
  scale_color_brewer(type = 'seq') +
  labs(title=
         "Comparison of pH and fixed acidity in terms of quality") + 
  labs(color = "Wine Quality")+ 
  geom_smooth(method = "lm",color = 'red',se = FALSE) + 
  theme_dark() + theme(legend.position = "bottom") + 
  xlab("pH") + ylab("Fixed Acidity")
```

> 
The quality of the wines can be seen above and below the line of
best fit in terms of fixed acidity and pH. The majority of the 
different qualities tend to be within 3.1 and 3.5 for the pH level 
and 6 and 10 g/dm3 for the fixed acidity. 

# Multivariate Analysis
> 
The multvariate plots with quality had some plots that were easily 
seen where the different scoring wines were within the plot. In terms
of pH, fixed acidity, and quality, the majority of wines that scored 
5 and 6 can be seen distributed above and below the line of best fit.
The citric acid, volatile acidity, and quality plot can be seen 
clearly where different quality of wines are congregating around a 
specific position below or above. In that plot, 6,7, and 8 scoring 
wines are seen below the line of best fit and 5  tends to be above 
the line. 

------

# Final Plots and Summary


### Plot One
```{r}
#quality
ggplot(data = redWine1, aes(x = as.factor(quality))) + 
  geom_bar() + ylab("Count") + 
  xlab("Quality (score between 0 and 10)") + 
  ggtitle("The quantity of wines based on quality") +  
  theme(plot.title = element_text(hjust = 0.5))

```

### Description One
> 
This plot was interesting to see that the majority of the wines 
got a score of five and six. I was highly surprised to not see 
many wines that could not manage to get over 7 as a score. 

### Plot Two
```{r}
ggplot(aes(factor(quality),
           alcohol),
           data = redWine1) +
  geom_jitter(alpha = .3) +
  geom_boxplot(alpha = .5,color = 'blue') +
  stat_summary(fun.y = "mean",
               geom = "point",
               color = 'red',
               shape = 8,
               size = 4) +
  labs(title="Comparison of quality and alcohol") +
  xlab("Quality (score between 0 and 10)") +
  ylab("Alcohol (% by volume)") +
  theme(plot.title = element_text(hjust = 0.5))
  


```

### Description Two
> 
The relation between quality and alcohol was surprising to see
a moderate positve correlation. 


### Plot Three
```{r}

ggplot(data = redWine1, aes(x=citric.acid,
                            y=volatile.acidity )) +
  geom_point(aes(color = as.factor(quality)))  + 
  scale_color_brewer(type = 'seq') + 
  labs(title=
"Comparison of citric acid and volatile acidity in terms of quality") + 
  labs(color = "Wine Quality")+ 
  geom_smooth(method = "lm",color = 'red',se = FALSE) +
  theme(legend.position = "bottom") + theme_dark() +
  xlab("Citric Acid (g/dm3) ") + 
  ylab("Volatile Acidity (g/dm3) ") +
  theme(plot.title = element_text(hjust = 0.5))
```

### Description Three
> 
Volatile acidity and citric acid plays a role with how great the
wine tastes.It is visible that the higher volatile acidity levels
tend to get scores of 5 and as you decrease the volatile acidity 
and increase the citric acid to about .30, you get to observe the 
majority of the 6 and 7's. This corresponds to the given 
documentation about volatile acidity and citric acid within this 
dataset.  

------

# Reflection

>  
While producing the univariate plots, many of the outputs had
noticeable positve skewness. When this occured, I would try to 
use the scale_x_log10 function to try to reduce the skewness. 
This did not work with alcohol, fixed acidity, volatile acidity,
citric acid, nor free sulfur dioxide. For volatile acidity, 
citric acid, and free sulfur dioxide it actually created a 
negative tail distribution. The improvement that the log10 
modification had on total sulfur dioxide, chlorides, residual sugars,
and sulphates was highly noticeable.The bivariate plot for alcohol 
and quality initially had a problem with overplotting and did not 
show great insight. This was improved by using geom_jitter to create
some random variation and an alpha value to help improve
the visibility of the scatterplot.The remaining bivariate plots also 
incorporated the alpha function to help produce better visibility 
of the individual data points.It was interesting to see the 
multivariate scatterplot between citric acid,volatile acidity, and 
quality. It was easily shown how the amount of volatile acidity and
citric acid plays a large role on the flavoring of the red wine and
its final score. Decreasing the volatile acidity and increasing the 
citric acid to about .30, gets the majority of the wines that scored
6 and 7's.A future project with this dataset would be to come up with
a predictive model that could calculate the quality of the wine by 
inputing the phsyiochemical values for a wine from the same 
geographical region. 
